Placeholder for article 3

### 2025-01-22 analysis/paper/paper.qmd
Final daily commit. Graph and table improvements. Results description improved and restructured (on-going). Document rendered.

Recent research has expanded our understanding of mobility and social networks through lithic raw material studies. Our team’s recent work (Belmiro et al., 2025) identified non-local cherts originating from central Portugal (>250 km from Vale Boi) and souuthern Spain (>200 km from Vale Boi). The proportions of these materials varied across the UP, with earlier occupations (Gravettian) showing high frequencies of non-local chert, while later ones (Proto-Solutrean and Solutrean) exhibited a marked decline. These findings suggest the establishment and maintenance of extensive social networks connecting southwestern Portugal to Southern spain and central Portugal as early as the beginning of the UP.

@@

The Terrace spans occupations from the UP to the Early Neolithic. Although excavations began in 2000, a new 8-square-meter area was opened in 2012 to better understand the stratigraphic sequence and assess the potential presence of older cultural layers, which could be linked to an early Upper Paleolithic occupation. Between 2012 and 2016, eight primary litho-stratigraphic units were identified. The UP sequence includes multiple occupations, spanning ~10,000 years: Gravettian (levels 8 to 6) between approximately c. 32 and 27 ka cal BP, Proto-Solutrean (levels 5 to 4E) between approximately c. 26 and 24 ka cal BP, and Solutrean (levels 4D, 4C, 4, and Lower 3) between approximately c. 24 and 20 ka cal BP (Belmiro et al., 2021; Cascalheira et al., 2017; Cascalheira & Bicho, 2013).

@@

The lithic analysis of the Shelter Solutrean levels (A-C) shows a high number of lithic artifacts, with over 22 000 artifacts analysed (including chips which correspond to ~60% of the assemblage and shatter ~17-19%). The volumes introduced into the site were mostly cortical, given the high frequencies of cortex in blanks and retouched tools. The knapping strategies are dominated by unidirectional and bidirectional prismatic, with frequent core maintenance products. Flakes dominate the debitage assemblage, and endscrapers, notches, splintered pieces and retouched blades/flakes are the most common retouched tools. Stemmed and winged points, laurel leaves, shouldered points and parpalló points were also identified (Cascalheira et al., 2013).

The Proto-Solutrean horizons yielded more than 26,500 artifacts (including chips which correspond to ~65-70% of the assemblage and shatter ~20%), ~11,000 attributed to the lower bottom of level 5 and ~15,600 attributed to the top of level 5 and level 4E. Both horizons are characterised by knapping strategies dominated by unidirectional prismatic cores, with little platform preparation, focused on the obtention of bladelets (in bottom level 5) and flakes. Retouched tools are dominated by splintered pieces, endscrapers and retouched blanks, with Vale Comprido points (the Proto-Solutrean fossil director) being barely inexistent in chert and mostly produced in dolerite (Belmiro et al., 2021).

Previous studies showed that the main raw materials used at the site are chert, quartz and greywacke, with the percentages of these changing between occupations. For example, while studies show a high degree of quartz in the lower Proto-Solutrean levels (Belmiro et al., 2021), in the Solutrean occupations of the Shelter a dominance of chert was observed (Cascalheira, 2010). Similarly, different raw materials have been observed to play specific roles, such as the case of chert for the production of a high percentage of blanks and retouched tools. While quartz and greywacke can be found locally, in the streams and adjacent to the site, chert sources are further away, although still within a ~20 km radius of the site. Our previous study showed, as expected, the existence of local chert in the assemblages, originating from the Lower Jurassic outcrops in proximity to the site (Belmiro et al., 2025). As previously mentioned, a portion of non-local cherts was also identified, with some identified sources, such as the Cretaceous outcrops of Central Portugal (Lisbon and Rio Maior areas) and the Upper Jurassic oolitic/peloidal cherts of the Betic Systems in southern Spain. This suggests contacts or long-range mobility patterns of the groups occupying Vale Boi all throughout the UP, although in different degrees, possibly related to changes in the settlement patterns and duration of site occupation.

@@

Using the Whole Assemblage Behavioral Index (WABI), previously used in other studies (Riel-Salvatore, 2010; Riel-Salvatore & Barton, 2004), Cascalheira et al., (2017) used the lithic results from the Shelter Solutrean levels (Cascalheira, 2010; Cascalheira et al., 2013) and Magdalenian levels (Mendonça, 2009), and the Gravettian levels of Terrace (Marreiros et al., 2015) to reconstruct and compare the mobility strategies of the several occupations, through the lithic volumetric density and the frequency of retouched tools. This method allowed the confirmation that, except from the Early Gravettian and Magdalenian occupations, all other occupations are composed of high-density assemblages with low degree of retouch, correspondent to a residential base-camp occupation, for extended periods of time. Regarding the Terrace area, however, the results were based on previous excavations (<2012), and the identification of new layers and reorganization of the occupations is yet to be tested through the use of lithic techno-typological methods.

A similar conclusion was suggested based on the interpretation of the chert sourcing and proportions in the several UP occupations of the Terrace and Shelter of Vale Boi, and following other models relating local/non-local raw materials with mobility (Grove et al., 2023; Surovell, 2009). The high proportions of non-local chert (~50%) in the Gravettian occupations (levels 6 and 7), and gradual reduction in the following occupations, were interpreted as a possible result of settlement changes, where the Gravettian occupations corresponded to short-term occupations, and the other occupations corresponded to residential, long-term occupations with more infrequent moves (Belmiro et al., 2025).

@@

As such, this study includes datasets which have been previously interpreted and published, most of which can also be found in online repositories and freely consulted. As previously mentioned, the macroscopic and petrographic data of cherts was previously collected and published by our team (Belmiro et al., 2025), and can also be found in an online repository (OSF**). This study focused on the UP chalcedony and chert assemblages of Vale Boi, especially the Gravettian (levels 6 and 7), Proto-Solutrean (levels 4E and 5) and Solutrean (levels 4 to 4D) from the Terrace area and the Solutrean (levels A to C) from the Shelter area. The methodology used in the study, and supporting documents, including the chert type characterisation and analysis data dictionaries can be found in the article and the online repository.

The technological analysis datasets have been collected by our team and published in previous works. The Solutrean assemblage technological analysis was published by Cascalheira (-Cascalheira, 2010), as part of a Master’s thesis and following publications, with the study of levels A to C of the Shelter area. The dataset used for this study was not originally published, but can be consulted in our online repository (++).  Given the inexistence of technological studies for the Solutrean levels of the Terrace with a representative amount of lithics (level 4), the Solutrean analysis in this study is limited to the Shelter area. However, and as seen in Belmiro et al. (2025), the raw material patterns seem to show little differences between the Solutrean in both areas, thus coupled with the similar radiocarbon dates between area (c. 24-22 and 24-20 ka cal BP), we assume that they are representative of similar, long-term, residential occupations which aptly characterise the Solutrean groups of Vale Boi. The Proto-Solutrean technological analysis was the result of a Master’s thesis (Belmiro, 2020) and following publication (Belmiro et al., 2021), with the study of levels 4E and 5 of the Terrace area. The dataset used for this study can be consulted in the original online repository (++), as well as in our RStudio reproducible project, available in our online repository (**).  Finally, the Gravettian technological analysis was the result of a recent and unpublished study, focusing on the study of levels 6 and 7 of the Terrace area. This dataset can also be found in our online repository (++). Given the use of different datasets, collected through a wide span of time and by different researchers, a data dictionary with the variables used in this study can also be found in the Supplementary materials and in our online repository.

@@

After data collection, the databases were imported into R Programming Environment, where the information was processed through the creation of descriptive statistical analysis and writing of this article. Some differences in the collection process, especially between the study by Cascalheira (2010) and the following studies, required homogenization. For example, where in Cascalheira (2010) retouch was an independent variable, with a Yes or No option followed by a variable to classify the type of retouch, in Belmiro (2020) and Belmiro et al. (2021), retouch was identified in the Class variable as RetouchedPiece, followed by a variable to choose the retouch typology. In this situation, the Solutrean dataset was corrected, so that all artifacts with retouch were classified as RetouchedPiece in the Class variable, instead of the original classification. The original datasets without alterations can be found in the online repository, as all the alterations to the original datasets were done through R Studio and can be found in the original R Script. Finally, technological and raw material data had to be merged in a single dataset. This was done through the merging of the several datasets through the artifact identifier “ID”, the sequential identifier which is attributed to each total station plot during excavation, whether relating to individual artifacts plotted in their exact location, or bucket coordinates, relating to the center of the excavated area, and attributed to all artifacts found in the sieved sediment in the bucket. Since the raw material study only focused piece plotted artifacts with individual coordinates (leaving out bucket artifacts), as these correspond to artifacts with dimensions superior to 2 cm or complete bladelets (thus excluding small fragments and chips), the final dataset used in this study does not directly correspond to the full datasets analysed in the original studies. Similarly, the raw material study used only currently accessible artifacts from the assemblages. This is especially relevant for the Shelter Solutrean chert assemblage, where shatter is absent from the sample due to the availability for study. This stands as a caveat for the current study, since it is using only portion of the assemblages, leaving out possibly important debris. Despite this, the current study still offers a novel and key approach to the study of lithic assemblages at the site, with a methodology that includes all the knapping sequence, of a large number of artifacts from the UP sequence of a single site, as seen in table **, which shows the number of artifacts used in the final analysis of this paper, after corrections, homogenization and merging.

After the homogenization and merging of the datasets, descriptive statistical analysis was applied to describe and interpret the relation between the raw material and technological data. As mentioned above, the goal of the study was to test two hypotheses related to the technological organisation of local and non-local cherts, and between different UP occupations. To test these hypotheses, we evaluated several parameters for each dataset and assemblage and compared the results between chert types and assemblages. The parameters evaluated in the current study were: P1) frequency of retouched tools; P2) frequency of maintenance products; P3) debitage sizes; P4) tools-to-debitage ratios; P5) core sizes; P6) blank-to-core ratios; P7) retouched tool sizes; P8) tool diversity; P9) tool specialization. Debitage, core and retouched tool sizes were analysed through descriptive statistics (mean, sd, min and max, tables available in the SOM**) and plotted through boxplots. The ratios were obtained through the division of each group (e.g., tools by debitage and blanks by core), by chert type and assemblage. All other parameters were observed through number and percentages tables. All the analyses were done through code in R Studio.

@@

# Tables with class by raw material type (n and %)
## For Gravettian
gravettian_general <- phase_dataset |>
  filter(Level %in% c(6, 7)) |>
  mutate(RMTYPE = factor(RMTYPE, levels = c("TYPE 1","TYPE 2", "TYPE 3", "TYPE 4",
                                        "TYPE 5", "TYPE 6","TYPE 7", 
                                        "TYPE 8", "TYPE 9", "TYPE 10", "TL", "INDET"))) |>
  tabyl(CLASS, RMTYPE) |> 
  adorn_totals(c("row", "col")) |> 
  adorn_percentages("col") |> 
  adorn_pct_formatting(2) |> 
  adorn_ns("front")

## For Proto-Solutrean (4E and top of 5)
proto_general <- phase_dataset |> 
  filter(Level %in% c("4E", "4E/5")) |> 
  filter(CLASS != "Chip") |> 
  mutate(RMTYPE = factor(RMTYPE, levels = c("TYPE 1","TYPE 2", "TYPE 3", "TYPE 4",
                                        "TYPE 5", "TYPE 6","TYPE 7", 
                                        "TYPE 8", "TYPE 9", "TYPE 10", "TL", "INDET"))) |> 
  tabyl(CLASS, RMTYPE) |> 
  adorn_totals(c("row", "col")) |> 
  adorn_percentages("col") |> 
  adorn_pct_formatting(2) |> 
  adorn_ns("front")

## For Solutrean
solutrean_general <- solutrean_dataset |> 
  mutate(RMTYPE = factor(RMTYPE, levels = c("TYPE 1","TYPE 2", "TYPE 3", "TYPE 4",
                                        "TYPE 5", "TYPE 6","TYPE 7", 
                                        "TYPE 8", "TYPE 9", "TYPE 10", "TYPE 11", "TL", "INDET"))) |> 
  tabyl(CLASS, RMTYPE) |> 
  adorn_totals(c("row", "col")) |> 
  adorn_percentages("col") |> 
  adorn_pct_formatting(2) |> 
  adorn_ns("front")

@@

# Scatterplot for width and length
solutrean_scatter <- solutrean_stats |>
  filter(CLASS %in% c("Flake","ElongatedProd", "Core", "RetouchedPiece")) |>
  filter(MAXWIDTH !="0" & LENGTH != "0" & !RMTYPE %in% c("INDET","GB","TL")) |>
  dplyr::mutate(RMTYPE = case_when(
    RMTYPE == "TYPE 1" ~ "T1",
    RMTYPE == "TYPE 2" ~ "T2",
    RMTYPE == "TYPE 3" ~ "T3",
    RMTYPE == "TYPE 4" ~ "T4",
    RMTYPE == "TYPE 5" ~ "T5",
    RMTYPE == "TYPE 6" ~ "T6",
    RMTYPE == "TYPE 7" ~ "T7",
    RMTYPE == "TYPE 8" ~ "T8",
    RMTYPE == "TYPE 9" ~ "T9",
    RMTYPE == "TYPE 10" ~ "T10",
    RMTYPE == "TYPE 11" ~ "T11",
    TRUE ~ RMTYPE)) |>
   mutate(RMTYPE = factor(RMTYPE, levels = c("T1", "T2", "T3", "T4", "T5", "T6","T7", 
                                        "T8", "T9", "T10", "T11"))) |>
  mutate(CLASS = factor(CLASS, levels = c("Flake", "ElongatedProd", "Core", "RetouchedPiece"))) |>
  ggplot(aes(x = LENGTH, y = MAXWIDTH, colour = RMTYPE)) +
  geom_point(alpha = 0.9, size = 3) +
  facet_wrap(~CLASS) +
  labs(x="Length (mm)", y="Max. Width (mm") +
  scale_colour_manual(values = c("#8B7B8B", "#FFE1FF", "#EED2EE", "#D8BFD8",
                              "#CDB5CD", "#2B4F60" , "#4A708B", "#798777",
                            "#9FB6CD", "#BDD2B6", "#BDC7C9"))

# Boxplot for individual measurements
# Weight
solutrean_stats_weight <- solutrean_stats |>
  filter(CLASS %in% c("Flake", "ElongatedProd", "Core", "RetouchedPiece")) |> 
  filter(WEIGHT !="0" & !RMTYPE %in% c("INDET","GB","TL")) |>
  mutate(RMTYPE = factor(RMTYPE, levels = c("TYPE 1","TYPE 2", "TYPE 3", "TYPE 4",
                                        "TYPE 5", "TYPE 6","TYPE 7", 
                                        "TYPE 8", "TYPE 9", "TYPE 10", "TYPE 11", "TL"))) |>
  mutate(CLASS = factor(CLASS, levels = c("Flake", "ElongatedProd", "Core", "RetouchedPiece"))) |>
  ggplot(aes(RMTYPE, WEIGHT, fill = RMTYPE)) +
  geom_boxplot(aes(fill=factor(RMTYPE)), outlier.shape=21) +
  facet_wrap(~CLASS) +
  stat_summary(fun.y=mean, geom="point", shape=23) +
  labs(x="Raw material types" ,y="Weight (gr)") +
  coord_cartesian(ylim =  c(0, 50)) +
  scale_fill_manual(values = c("#8B7B8B", "#FFE1FF", "#EED2EE", "#D8BFD8",
                              "#CDB5CD", "#2B4F60" , "#4A708B", "#798777",
                            "#9FB6CD", "#BDD2B6", "#BDC7C9")) +
  theme(axis.text.x=element_blank(),
        axis.ticks.x=element_blank())

@@

# Scatterplots for width and length
proto_scatter <- proto_stats |>
  filter(CLASS %in% c("Flake","ElongatedProd", "Core", "RetouchedPiece")) |>
  filter(MAXWIDTH !="0" & LENGTH != "0" & !RMTYPE %in% c("INDET","GB","TL")) |>
  dplyr::mutate(RMTYPE = case_when(
    RMTYPE == "TYPE 1" ~ "T1",
    RMTYPE == "TYPE 2" ~ "T2",
    RMTYPE == "TYPE 3" ~ "T3",
    RMTYPE == "TYPE 4" ~ "T4",
    RMTYPE == "TYPE 5" ~ "T5",
    RMTYPE == "TYPE 6" ~ "T6",
    RMTYPE == "TYPE 7" ~ "T7",
    RMTYPE == "TYPE 8" ~ "T8",
    RMTYPE == "TYPE 9" ~ "T9",
    RMTYPE == "TYPE 10" ~ "T10",
    TRUE ~ RMTYPE)) |>
   mutate(RMTYPE = factor(RMTYPE, levels = c("T1", "T2", "T3", "T4", "T5", "T6","T7", 
                                        "T8", "T9", "T10"))) |>
  mutate(CLASS = factor(CLASS, levels = c("Flake", "ElongatedProd", "Core", "RetouchedPiece"))) |>
  ggplot(aes(x = LENGTH, y = MAXWIDTH, colour = RMTYPE)) +
  geom_point(alpha = 0.9, size = 3) +
  facet_wrap(~CLASS) +
  labs(x="Length (mm)", y="Max. Width (mm") +
  scale_colour_manual(values = c("#8B7B8B", "#FFE1FF", "#EED2EE", "#D8BFD8",
                              "#CDB5CD", "#2B4F60" , "#4A708B", "#798777",
                            "#9FB6CD", "#BDD2B6"))

>>>>

Recent research has expanded our understanding of mobility and social networks through lithic raw material studies. Our team’s recent work (Belmiro et al., 2025) identified non-local cherts originating from central Portugal (\>250 km from Vale Boi) and souuthern Spain (\>200 km from Vale Boi). The proportions of these materials varied across the UP, with earlier occupations (Gravettian) showing high frequencies of non-local chert, while later ones (Proto-Solutrean and Solutrean) exhibited a marked decline. These findings suggest the establishment and maintenance of extensive social networks connecting southwestern Portugal to Southern spain and central Portugal as early as the beginning of the UP.

@@

The Terrace spans occupations from the UP to the Early Neolithic. Although excavations began in 2000, a new 8-square-meter area was opened in 2012 to better understand the stratigraphic sequence and assess the potential presence of older cultural layers, which could be linked to an early Upper Paleolithic occupation. Between 2012 and 2016, eight primary litho-stratigraphic units were identified. The UP sequence includes multiple occupations, spanning \~10,000 years: Gravettian (levels 8 to 6) between approximately c. 32 and 27 ka cal BP, Proto-Solutrean (levels 5 to 4E) between approximately c. 26 and 24 ka cal BP, and Solutrean (levels 4D, 4C, 4, and Lower 3) between approximately c. 24 and 20 ka cal BP (Belmiro et al., 2021; Cascalheira et al., 2017; Cascalheira & Bicho, 2013).

@@

The lithic analysis of the Shelter Solutrean levels (A-C) shows a high number of lithic artifacts, with over 22 000 artifacts analysed (including chips which correspond to \~60% of the assemblage and shatter \~17-19%). The volumes introduced into the site were mostly cortical, given the high frequencies of cortex in blanks and retouched tools. The knapping strategies are dominated by unidirectional and bidirectional prismatic, with frequent core maintenance products. Flakes dominate the debitage assemblage, and endscrapers, notches, splintered pieces and retouched blades/flakes are the most common retouched tools. Stemmed and winged points, laurel leaves, shouldered points and parpalló points were also identified (Cascalheira et al., 2013).

The Proto-Solutrean horizons yielded more than 26,500 artifacts (including chips which correspond to \~65-70% of the assemblage and shatter \~20%), \~11,000 attributed to the lower bottom of level 5 and \~15,600 attributed to the top of level 5 and level 4E. Both horizons are characterised by knapping strategies dominated by unidirectional prismatic cores, with little platform preparation, focused on the obtention of bladelets (in bottom level 5) and flakes. Retouched tools are dominated by splintered pieces, endscrapers and retouched blanks, with Vale Comprido points (the Proto-Solutrean fossil director) being barely inexistent in chert and mostly produced in dolerite (Belmiro et al., 2021).

Previous studies showed that the main raw materials used at the site are chert, quartz and greywacke, with the percentages of these changing between occupations. For example, while studies show a high degree of quartz in the lower Proto-Solutrean levels (Belmiro et al., 2021), in the Solutrean occupations of the Shelter a dominance of chert was observed (Cascalheira, 2010). Similarly, different raw materials have been observed to play specific roles, such as the case of chert for the production of a high percentage of blanks and retouched tools. While quartz and greywacke can be found locally, in the streams and adjacent to the site, chert sources are further away, although still within a \~20 km radius of the site. Our previous study showed, as expected, the existence of local chert in the assemblages, originating from the Lower Jurassic outcrops in proximity to the site (Belmiro et al., 2025). As previously mentioned, a portion of non-local cherts was also identified, with some identified sources, such as the Cretaceous outcrops of Central Portugal (Lisbon and Rio Maior areas) and the Upper Jurassic oolitic/peloidal cherts of the Betic Systems in southern Spain. This suggests contacts or long-range mobility patterns of the groups occupying Vale Boi all throughout the UP, although in different degrees, possibly related to changes in the settlement patterns and duration of site occupation.

@@

Using the Whole Assemblage Behavioral Index (WABI), previously used in other studies (Riel-Salvatore, 2010; Riel-Salvatore & Barton, 2004), Cascalheira et al., (2017) used the lithic results from the Shelter Solutrean levels (Cascalheira, 2010; Cascalheira et al., 2013) and Magdalenian levels (Mendonça, 2009), and the Gravettian levels of Terrace (Marreiros et al., 2015) to reconstruct and compare the mobility strategies of the several occupations, through the lithic volumetric density and the frequency of retouched tools. This method allowed the confirmation that, except from the Early Gravettian and Magdalenian occupations, all other occupations are composed of high-density assemblages with low degree of retouch, correspondent to a residential base-camp occupation, for extended periods of time. Regarding the Terrace area, however, the results were based on previous excavations (\<2012), and the identification of new layers and reorganization of the occupations is yet to be tested through the use of lithic techno-typological methods.

A similar conclusion was suggested based on the interpretation of the chert sourcing and proportions in the several UP occupations of the Terrace and Shelter of Vale Boi, and following other models relating local/non-local raw materials with mobility (Grove et al., 2023; Surovell, 2009). The high proportions of non-local chert (\~50%) in the Gravettian occupations (levels 6 and 7), and gradual reduction in the following occupations, were interpreted as a possible result of settlement changes, where the Gravettian occupations corresponded to short-term occupations, and the other occupations corresponded to residential, long-term occupations with more infrequent moves (Belmiro et al., 2025).

@@

As such, this study includes datasets which have been previously interpreted and published, most of which can also be found in online repositories and freely consulted. As previously mentioned, the macroscopic and petrographic data of cherts was previously collected and published by our team (Belmiro et al., 2025), and can also be found in an online repository (OSF\*\*). This study focused on the UP chalcedony and chert assemblages of Vale Boi, especially the Gravettian (levels 6 and 7), Proto-Solutrean (levels 4E and 5) and Solutrean (levels 4 to 4D) from the Terrace area and the Solutrean (levels A to C) from the Shelter area. The methodology used in the study, and supporting documents, including the chert type characterisation and analysis data dictionaries can be found in the article and the online repository.

The technological analysis datasets have been collected by our team and published in previous works. The Solutrean assemblage technological analysis was published by Cascalheira (-Cascalheira, 2010), as part of a Master’s thesis and following publications, with the study of levels A to C of the Shelter area. The dataset used for this study was not originally published, but can be consulted in our online repository (++). Given the inexistence of technological studies for the Solutrean levels of the Terrace with a representative amount of lithics (level 4), the Solutrean analysis in this study is limited to the Shelter area. However, and as seen in Belmiro et al. (2025), the raw material patterns seem to show little differences between the Solutrean in both areas, thus coupled with the similar radiocarbon dates between area (c. 24-22 and 24-20 ka cal BP), we assume that they are representative of similar, long-term, residential occupations which aptly characterise the Solutrean groups of Vale Boi. The Proto-Solutrean technological analysis was the result of a Master’s thesis (Belmiro, 2020) and following publication (Belmiro et al., 2021), with the study of levels 4E and 5 of the Terrace area. The dataset used for this study can be consulted in the original online repository (++), as well as in our RStudio reproducible project, available in our online repository (\*\*). Finally, the Gravettian technological analysis was the result of a recent and unpublished study, focusing on the study of levels 6 and 7 of the Terrace area. This dataset can also be found in our online repository (++). Given the use of different datasets, collected through a wide span of time and by different researchers, a data dictionary with the variables used in this study can also be found in the Supplementary materials and in our online repository.

@@

After data collection, the databases were imported into R Programming Environment, where the information was processed through the creation of descriptive statistical analysis and writing of this article. Some differences in the collection process, especially between the study by Cascalheira (2010) and the following studies, required homogenization. For example, where in Cascalheira (2010) retouch was an independent variable, with a Yes or No option followed by a variable to classify the type of retouch, in Belmiro (2020) and Belmiro et al. (2021), retouch was identified in the Class variable as RetouchedPiece, followed by a variable to choose the retouch typology. In this situation, the Solutrean dataset was corrected, so that all artifacts with retouch were classified as RetouchedPiece in the Class variable, instead of the original classification. The original datasets without alterations can be found in the online repository, as all the alterations to the original datasets were done through R Studio and can be found in the original R Script. Finally, technological and raw material data had to be merged in a single dataset. This was done through the merging of the several datasets through the artifact identifier “ID”, the sequential identifier which is attributed to each total station plot during excavation, whether relating to individual artifacts plotted in their exact location, or bucket coordinates, relating to the center of the excavated area, and attributed to all artifacts found in the sieved sediment in the bucket. Since the raw material study only focused piece plotted artifacts with individual coordinates (leaving out bucket artifacts), as these correspond to artifacts with dimensions superior to 2 cm or complete bladelets (thus excluding small fragments and chips), the final dataset used in this study does not directly correspond to the full datasets analysed in the original studies. Similarly, the raw material study used only currently accessible artifacts from the assemblages. This is especially relevant for the Shelter Solutrean chert assemblage, where shatter is absent from the sample due to the availability for study. This stands as a caveat for the current study, since it is using only portion of the assemblages, leaving out possibly important debris. Despite this, the current study still offers a novel and key approach to the study of lithic assemblages at the site, with a methodology that includes all the knapping sequence, of a large number of artifacts from the UP sequence of a single site, as seen in table \*\*, which shows the number of artifacts used in the final analysis of this paper, after corrections, homogenization and merging.

After the homogenization and merging of the datasets, descriptive statistical analysis was applied to describe and interpret the relation between the raw material and technological data. As mentioned above, the goal of the study was to test two hypotheses related to the technological organisation of local and non-local cherts, and between different UP occupations. To test these hypotheses, we evaluated several parameters for each dataset and assemblage and compared the results between chert types and assemblages. The parameters evaluated in the current study were: P1) frequency of retouched tools; P2) frequency of maintenance products; P3) debitage sizes; P4) tools-to-debitage ratios; P5) core sizes; P6) blank-to-core ratios; P7) retouched tool sizes; P8) tool diversity; P9) tool specialization. Debitage, core and retouched tool sizes were analysed through descriptive statistics (mean, sd, min and max, tables available in the SOM\*\*) and plotted through boxplots. The ratios were obtained through the division of each group (e.g., tools by debitage and blanks by core), by chert type and assemblage. All other parameters were observed through number and percentages tables. All the analyses were done through code in R Studio.

@@

# Tables with class by raw material type (n and %)
## For Gravettian
gravettian_general <- phase_dataset |>
  filter(Level %in% c(6, 7)) |>
  mutate(RMTYPE = factor(RMTYPE, levels = c("TYPE 1","TYPE 2", "TYPE 3", "TYPE 4",
                                        "TYPE 5", "TYPE 6","TYPE 7", 
                                        "TYPE 8", "TYPE 9", "TYPE 10", "TL", "INDET"))) |>
  tabyl(CLASS, RMTYPE) |> 
  adorn_totals(c("row", "col")) |> 
  adorn_percentages("col") |> 
  adorn_pct_formatting(2) |> 
  adorn_ns("front") |> 
  mutate(across(where(is.character), ~ gsub("^0   \\(0\\.00%\\)$", " ", .)))

## For Proto-Solutrean (4E and top of 5)
proto_general <- phase_dataset |> 
  filter(Level %in% c("4E", "4E/5")) |> 
  filter(CLASS != "Chip") |> 
  mutate(RMTYPE = factor(RMTYPE, levels = c("TYPE 1","TYPE 2", "TYPE 3", "TYPE 4",
                                        "TYPE 5", "TYPE 6","TYPE 7", 
                                        "TYPE 8", "TYPE 9", "TYPE 10", "TL", "INDET"))) |> 
  tabyl(CLASS, RMTYPE) |> 
  adorn_totals(c("row", "col")) |> 
  adorn_percentages("col") |> 
  adorn_pct_formatting(2) |> 
  adorn_ns("front") |> 
  mutate(across(where(is.character), ~ gsub("^0   \\(0\\.00%\\)$", " ", .)))

## For Solutrean
solutrean_general <- solutrean_dataset |> 
  mutate(RMTYPE = factor(RMTYPE, levels = c("TYPE 1","TYPE 2", "TYPE 3", "TYPE 4",
                                        "TYPE 5", "TYPE 6","TYPE 7", 
                                        "TYPE 8", "TYPE 9", "TYPE 10", "TYPE 11", "TL", "INDET"))) |> 
  tabyl(CLASS, RMTYPE) |> 
  adorn_totals(c("row", "col")) |> 
  adorn_percentages("col") |> 
  adorn_pct_formatting(2) |> 
  adorn_ns("front") |> 
  mutate(across(where(is.character), ~ gsub("^0   \\(0\\.00%\\)$", " ", .)))

@@

# Scatterplot for width and length
solutrean_scatter <- solutrean_stats |>
  filter(CLASS %in% c("Flake","ElongatedProd", "Core", "RetouchedPiece")) |>
  filter(MAXWIDTH !="0" & LENGTH != "0" & !RMTYPE %in% c("INDET","GB","TL")) |>
  dplyr::mutate(RMTYPE = case_when(
    RMTYPE == "TYPE 1" ~ "T1",
    RMTYPE == "TYPE 2" ~ "T2",
    RMTYPE == "TYPE 3" ~ "T3",
    RMTYPE == "TYPE 4" ~ "T4",
    RMTYPE == "TYPE 5" ~ "T5",
    RMTYPE == "TYPE 6" ~ "T6",
    RMTYPE == "TYPE 7" ~ "T7",
    RMTYPE == "TYPE 8" ~ "T8",
    RMTYPE == "TYPE 9" ~ "T9",
    RMTYPE == "TYPE 10" ~ "T10",
    RMTYPE == "TYPE 11" ~ "T11",
    TRUE ~ RMTYPE)) |>
   mutate(RMTYPE = factor(RMTYPE, levels = c("T1", "T2", "T3", "T4", "T5", "T6","T7", 
                                        "T8", "T9", "T10", "T11"))) |>
  mutate(CLASS = factor(CLASS, levels = c("Flake", "ElongatedProd", "Core", "RetouchedPiece"))) |>
  ggplot(aes(x = LENGTH, y = MAXWIDTH, fill = RMTYPE)) +
  geom_point(alpha = 0.9, size = 3, shape = 21, colour = "grey") +
  facet_wrap(~CLASS) +
  labs(x="Length (mm)", y="Max. Width (mm") +
  scale_fill_manual(values = c("#8B7B8B", "#FFE1FF", "#EED2EE", "#D8BFD8",
                              "#CDB5CD", "#2B4F60" , "#4A708B", "#798777",
                            "#9FB6CD", "#BDD2B6", "#BDC7C9"))

# Boxplot for individual measurements
# Weight
solutrean_stats_weight <- solutrean_stats |>
  filter(CLASS %in% c("Flake", "ElongatedProd", "Core", "RetouchedPiece")) |> 
  filter(WEIGHT !="0" & !RMTYPE %in% c("INDET","GB","TL")) |>
  mutate(RMTYPE = factor(RMTYPE, levels = c("TYPE 1","TYPE 2", "TYPE 3", "TYPE 4",
                                        "TYPE 5", "TYPE 6","TYPE 7", 
                                        "TYPE 8", "TYPE 9", "TYPE 10", "TYPE 11", "TL"))) |>
  mutate(CLASS = factor(CLASS, levels = c("Flake", "ElongatedProd", "Core", "RetouchedPiece"))) |>
  ggplot(aes(RMTYPE, WEIGHT, fill = RMTYPE)) +
  geom_boxplot(aes(fill=factor(RMTYPE)), outlier.shape=21) +
  facet_wrap(~CLASS, scales = "free_y") +
  stat_summary(fun.y=mean, geom="point", shape=23) +
  labs(x="Raw material types" ,y="Weight (gr)") +
  #coord_cartesian(ylim =  c(0, 50)) +
  scale_fill_manual(values = c("#8B7B8B", "#FFE1FF", "#EED2EE", "#D8BFD8",
                              "#CDB5CD", "#2B4F60" , "#4A708B", "#798777",
                            "#9FB6CD", "#BDD2B6", "#BDC7C9")) +
  theme(axis.text.x=element_blank(),
        axis.ticks.x=element_blank())

@@

# Scatterplots for width and length
proto_scatter <- proto_stats |>
  filter(CLASS %in% c("Flake","ElongatedProd", "Core", "RetouchedPiece")) |>
  filter(MAXWIDTH !="0" & LENGTH != "0" & !RMTYPE %in% c("INDET","GB","TL")) |>
  dplyr::mutate(RMTYPE = case_when(
    RMTYPE == "TYPE 1" ~ "T1",
    RMTYPE == "TYPE 2" ~ "T2",
    RMTYPE == "TYPE 3" ~ "T3",
    RMTYPE == "TYPE 4" ~ "T4",
    RMTYPE == "TYPE 5" ~ "T5",
    RMTYPE == "TYPE 6" ~ "T6",
    RMTYPE == "TYPE 7" ~ "T7",
    RMTYPE == "TYPE 8" ~ "T8",
    RMTYPE == "TYPE 9" ~ "T9",
    RMTYPE == "TYPE 10" ~ "T10",
    TRUE ~ RMTYPE)) |>
   mutate(RMTYPE = factor(RMTYPE, levels = c("T1", "T2", "T3", "T4", "T5", "T6","T7", 
                                        "T8", "T9", "T10"))) |>
  mutate(CLASS = factor(CLASS, levels = c("Flake", "ElongatedProd", "Core", "RetouchedPiece"))) |>
  ggplot(aes(x = LENGTH, y = MAXWIDTH, fill = RMTYPE)) +
  geom_point(alpha = 0.9, size = 3, shape = 21, colour = "grey") +
  facet_wrap(~CLASS) +
  labs(x="Length (mm)", y="Max. Width (mm") +
  scale_fill_manual(values = c("#8B7B8B", "#FFE1FF", "#EED2EE", "#D8BFD8",
                              "#CDB5CD", "#2B4F60" , "#4A708B", "#798777",
                            "#9FB6CD", "#BDD2B6"))

# Boxplots weight
proto_stats_weight <- proto_stats |>
  filter(CLASS %in% c("Flake","ElongatedProd", "Core", "RetouchedPiece")) |> 
  filter(WEIGHT !="0" & !RMTYPE %in% c("INDET","GB","TL")) |>
  mutate(RMTYPE = factor(RMTYPE, levels = c("TYPE 1","TYPE 2", "TYPE 3", "TYPE 4",
                                        "TYPE 5", "TYPE 6","TYPE 7", 
                                        "TYPE 8", "TYPE 9", "TYPE 10"))) |>
  mutate(CLASS = factor(CLASS, levels = c("Flake", "ElongatedProd", "Core", "RetouchedPiece"))) |>
  ggplot(aes(RMTYPE, WEIGHT, fill = RMTYPE)) +
  geom_boxplot(aes(fill=factor(RMTYPE)), outlier.shape=21) +
  facet_wrap(~CLASS, scales = "free_y") +
  stat_summary(fun.y=mean, geom="point", shape=23) +
  labs(x="Raw material types", y="Weight (gr)") +
  #coord_cartesian(ylim =  c(0, 50)) +
  scale_fill_manual(values = c("#8B7B8B", "#FFE1FF", "#EED2EE", "#D8BFD8",
                              "#CDB5CD", "#2B4F60" , "#4A708B", "#798777",
                            "#9FB6CD", "#BDD2B6")) +
  theme(axis.text.x=element_blank(),
        axis.ticks.x=element_blank())
